---
title: "Practical Machine Learning Project"
author: "Paolo Baglioni"
date: "23 March 2016"
output: html_document
---

## Background and Introduction
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. This project will use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. We were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

The goal of this assignment is to predict the manner in which they did the exercise. This assignment will:

a. create a report describing how the model is built
b. how cross validation is use
c. what is the expected out of sample error and why we made the choices we did
d. use prediction model to predict 20 different test cases 

## Data Processing

### Import the data
First load the R packages needed for analysis and then download the training and testing data sets from the given URLs.
```{r, message=FALSE}
# load the required packages
library(caret)
library(rattle)
library(rpart)
library(rpart.plot)
library(randomForest)
library(repmis)
```

```{r, message=FALSE}
# import the data from the URLs
# trainurl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
# testurl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
# training <- source_data(trainurl, na.strings = c("NA", "#DIV/0!", ""), header = TRUE)
# testing <- source_data(testurl, na.strings = c("NA", "#DIV/0!", ""), header = TRUE)
# load data locally
# setwd("C:/Users/baglioni/Desktop/Practical Machine Learning Project/data")
training <- read.csv("pml-training.csv", na.strings = c("NA", ""))
testing <- read.csv("pml-testing.csv", na.strings = c("NA", ""))
```


The training dataset has `r dim(training)[1]` observations and `r dim(training)[2]` variables, and the testing data set contains `r dim(testing)[1]` observations and the same variables as the training set. We are trying to predict the outcome of the variable `classe` in the training set.

### Data cleaning
Delete columns (predictors) of the training set that contain any missing values. 
```{r}
training <- training[, colSums(is.na(training)) == 0]
testing <- testing[, colSums(is.na(testing)) == 0]
```
Remove the first seven predictors since these variables have little predicting power for the outcome `classe`.
```{r}
trainData <- training[, -c(1:7)]
testData <- testing[, -c(1:7)]
```
The cleaned data sets `trainData` and `testData` both have 53 columns with the same first 52 variables and the last variable `classe` and `problem_id` individually. `trainData` has 19622 rows while `testData` has 20 rows.

### Data Partitioning

In order to get out-of-sample errors, we split the cleaned training set `trainData` into a training set (`train`, 70%) for prediction and a validation set (`valid` 30%) to compute the out-of-sample errors.
```{r}
set.seed(7826) 
inTrain <- createDataPartition(trainData$classe, p = 0.7, list = FALSE)
train <- trainData[inTrain, ]
valid <- trainData[-inTrain, ]
```


## Prediction Algorithms

Use classification trees and random forests to predict the outcome. 

### Classification trees
In practice, $k = 5$ or $k = 10$ when doing k-fold cross validation. Here we consider 5-fold cross validation (default setting in trainControl function is 10) when implementing the algorithm to save a little computing time. Since data transformations may be less important in non-linear models like classification trees, we do not transform any variables. 
```{r}
control <- trainControl(method = "cv", number = 5)
fit_rpart <- train(classe ~ ., data = train, method = "rpart", 
                   trControl = control)
print(fit_rpart, digits = 4)
fancyRpartPlot(fit_rpart$finalModel)
# predict outcomes using validation set
predict_rpart <- predict(fit_rpart, valid)
# Show prediction result
(conf_rpart <- confusionMatrix(valid$classe, predict_rpart))
(accuracy_rpart <- conf_rpart$overall[1])
```
From the confusion matrix, the accuracy rate is `r round(accuracy_rpart, 3)`, and so the out-of-sample error rate is `r 1 - round(accuracy_rpart, 3)`. Using classification tree does not predict the outcome `classe` very well.

### Random forests
Since classification tree method does not perform well, we try random forest method instead.
```{r}
fit_rf <- train(classe ~ ., data = train, method = "rf", 
                   trControl = control)
print(fit_rf, digits = 4)
# predict outcomes using validation set
predict_rf <- predict(fit_rf, valid)
# Show prediction result
(conf_rf <- confusionMatrix(valid$classe, predict_rf))
(accuracy_rf <- conf_rf$overall[1])
```
For this dataset, random forest is way better result as it has a higher accuracy of 0.9959, as compared to Decision Tree model, which has an accuracy of 0.7385. The Random Forest model is choosen.
The expected out-of-sample error is estimated at 0.005, or 0.5% accuracy for predictions made against the cross-validation set. 

## Prediction on Testing Set
Use random forests to predict the outcome variable `classe` for the testing set. 
```{r}
(predict(fit_rf, testData))
```


